### 2025-03-02 sec/1_introduction.tex
more minor fixes

\section{Introduction}
\label{sec:introduction}
Single cell detection in microscopy images is a fundamental task in biological and medical applications, enabling quantitative analysis of cellular processes and disease mechanisms~\cite{meijering_cell_2012}.
Accurate identification and localization of individual cells are crucial for understanding cellular heterogeneity, dynamics, and responses to stimuli, driving advancements in fields ranging from drug discovery to diagnostics~\cite{f_mualla_automatic_2016,e_d_ferreira_classification_2024}.
Brightfield microscopy, a label-free technique, is widely used due to its non-invasiveness and simplicity, allowing for long-term observation of live cells without phototoxicity or alteration of cellular states~\cite{jyrki_selinummi_bright_2009,huixia_ren_cellbow_2020}.
However, brightfield microscopy presents significant challenges for automated analysis due to inherent limitations such as low contract between cells and background, high variability in cell appearance influenced by factors like cell type and imaging conditions, and the presence of artifacts like uneven illumination and well edge distortions~\cite{jyrki_selinummi_bright_2009,f_mualla_automatic_2016}.
These challenges often necessitate labor-intensive manual annotation, hindering high-throughput analysis and objective analysis.

@@

This paper explores the use of unconditional diffusion models for generating synthetic brightfield microscopy images of CHO cells and investigates their impact on single cell detection accuracy using state-of-the-art object detectors.
The diffusion model was trained on a carefully curated dataset of 10,000 images (512x512 pixels) of stably transfected CHO-K1 and CHO DG44 cell lines, acquired using a high-throughput microscope.
We address the central research question: \textbf{Can synthetic brightfield microscopy images generated by diffusion models enhance the performance of object detection models for single cell detection?}
We further investigate: (1) the perceptual realism of generated images through expert evaluation, and (2) the influence of synthetic data proportion in training datasets.

>>>>

\section{Introduction}
\label{sec:introduction}
Single cell detection in microscopy images is a fundamental task in biological and medical applications, enabling quantitative analysis of cellular processes and disease mechanisms~\cite{meijering_cell_2012}.
Accurate identification and localization of individual cells are crucial for understanding cellular heterogeneity, dynamics, and responses to stimuli, driving advancements in fields ranging from drug discovery to diagnostics~\cite{f_mualla_automatic_2016,e_d_ferreira_classification_2024}.
Brightfield microscopy, a label-free technique, is widely used due to its non-invasiveness and simplicity, allowing for long-term observation of live cells without phototoxicity or alteration of cellular states~\cite{jyrki_selinummi_bright_2009,huixia_ren_cellbow_2020}.
However, brightfield microscopy presents significant challenges for automated analysis due to inherent limitations such as low contrast between cells and background, high variability in cell appearance influenced by factors like cell type and imaging conditions, and the presence of artifacts like uneven illumination and well edge distortions~\cite{jyrki_selinummi_bright_2009,f_mualla_automatic_2016}.
These challenges often necessitate labor-intensive manual annotation, hindering high-throughput analysis and objective analysis.

@@

This paper explores the use of unconditional diffusion models for generating synthetic brightfield microscopy images of CHO cells and investigates their impact on single cell detection accuracy using state-of-the-art object detectors.
The diffusion model was trained on a carefully curated dataset of 10,000 images (512x512 pixels) of stably transfected CHO-K1 and CHO DG44 cell lines, acquired using a high-throughput microscope.
We address the central research question: \textbf{Can synthetic brightfield microscopy images generated by diffusion models be used to enhance the performance of object detection models for single cell detection?}
We further investigate: (1) the perceptual realism of generated images through expert evaluation, and (2) the influence of synthetic data proportion in training datasets.

### 2025-03-02 sec/3_methodology.tex
more minor fixes

For the final synthetic image generation, the Euler a scheduler was chosen for its balance of sample quality and generation speed, determined through empirical testing of various schedulers.
Trailing timestep spacing and epsilon prediction type were used based on hyperparameter optimization experiments that indicated these settings produced visually superior results for our dataset.
To introduce variability and potentially enhance the robustness of the downstream object detectors, the number of inference steps during the final generation process was randomized between 35 and 40 steps for each batch of 16 images.
A total of 10,000 synthetic images were generated using this optimized configuration to create datasets for object detection model training.

@@

Synthetic images were labeled using a model-assisted labeling approach.
A YOLOv8m model was initially pre-trained and then fine-tuned on a subset of real labeled images.
This fine-tuned YOLOv8m model was then integrated into the Roboflow annotation tool's ``Model-Assisted Labeling'' feature.
These model-predicted bounding boxes were then manually reviewed and refined by a human expert to correct any inaccuracies, missed detections or false positives.
This approach leveraged the pre-trained object detection model to expedite the labeling of synthetic data, while manual refinement ensured the final labels were of high quality.

>>>>

For the final synthetic image generation, the Euler Ancestral scheduler was chosen for its balance of sample quality and generation speed, determined through empirical testing of various schedulers.
Trailing timestep spacing and epsilon prediction type were used based on hyperparameter optimization experiments that indicated these settings produced visually superior results for our dataset.
To introduce variability and potentially enhance the robustness of the downstream object detectors, the number of inference steps during the final generation process was randomized between 35 and 40 steps for each batch of 16 images.
A total of 10,000 synthetic images were generated using this optimized configuration to create datasets for object detection model training.

@@

Synthetic images were labeled using a model-assisted labeling approach.
A pretrained YOLOv8m model was fine-tuned on a subset of real labeled images.
This fine-tuned YOLOv8m model was then integrated into the Roboflow annotation tool's ``Model-Assisted Labeling'' feature.
These model-predicted bounding boxes were then manually reviewed and refined by a human expert to correct any inaccuracies, missed detections or false positives.
This approach leveraged the pre-trained object detection model to expedite the labeling of synthetic data, while manual refinement ensured the final labels were of high quality.

### 2025-03-02 sec/5_discussion.tex
more minor fixes

The implications of our findings for biological research and applications are two-fold.
The demonstrated ability to generate high-quality synthetic brightfield microscopy images and effectively utilize them for training cell detection  models opens up new avenues for addressing data scarcity and annotation bottlenecks in microscopy image analysis.
Synthetic data generation offers a cost-effective and scalable approach to augment limited real datasets, potentially democratizing access to advanced cell detection techniques, particularly for research groups with limited biological resources or access to large annotated datasets.
Furthermore, the use of synthetic data can reduce the dependence on time-consuming and expensive manual labor work, saving time and resources researcher usually have to invest in the laboratorial process.
The potential to reduce or eliminate the need for fluorescent labels in certain applications, by relying on robust detection models trained with label-free brightfield images augmented by synthetic data, also aligns with ethical considerations and best practices in cell biology, minimizing potential cytotoxic effects and enabling more physiologically relevant live-cell imaging studies.
Moreover, synthetic data generation offers the unique capability to create diverse datasets encompassing rare cell phenotypes or challenging imaging conditions that are difficult to obtain in real-world experiments, potentially enhancing the robustness nad generalizability of cell detection models for a wider range of biological applications and scenarios.
Transparency in reporting the use of synthetic data in research is crucial for maintaining scientific rigor and ethical standards, ensuring appropriate interpretation and validation of findings derived from models trained with synthetic data augmentation.

>>>>

The implications of our findings for biological research and applications are two-fold.
The demonstrated ability to generate high-quality synthetic brightfield microscopy images and effectively utilize them for training cell detection  models opens up new avenues for addressing data scarcity and annotation bottlenecks in microscopy image analysis.
Synthetic data generation offers a cost-effective and scalable approach to augment limited real datasets, potentially democratizing access to advanced cell detection techniques, particularly for research groups with limited biological resources or access to large annotated datasets.
Furthermore, the use of synthetic data can reduce the dependence on time-consuming and expensive manual labor work, saving time and resources researchers usually have to invest in the laboratorial process.
The potential to reduce or eliminate the need for fluorescent labels in certain applications, by relying on robust detection models trained with label-free brightfield images augmented by synthetic data, also aligns with ethical considerations and best practices in cell biology, minimizing potential cytotoxic effects and enabling more physiologically relevant live-cell imaging studies.
Moreover, synthetic data generation offers the unique capability to create diverse datasets encompassing rare cell phenotypes or challenging imaging conditions that are difficult to obtain in real-world experiments, potentially enhancing the robustness and generalizability of cell detection models for a wider range of biological applications and scenarios.
Transparency in reporting the use of synthetic data in research is crucial for maintaining scientific rigor and ethical standards, ensuring appropriate interpretation and validation of findings derived from models trained with synthetic data augmentation.
